---
title: "Ridge & Lasso Regression"
output:
      html_document:
        keep_md: true
date: "2026-01-06"
---

# Ridge and lasso regression with bootstrap

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidymodels)
library(sjPlot)
library(psych)
library(parallel)
library(finalfit)
library(gtsummary)
library(mlbench)
library(vip)
library(rsample)
library(tune)
library(recipes)
library(yardstick)
library(parsnip)
library(glmnet)
library(themis)
library(corrr)
library(performance)
library(utils)

data <- read_csv("canpath_imputed.csv")
```

## Linear Regression

A linear regression is a type of regression where the outcome variable is continuous and we assume it has a normal distribution. 

#### Feature selection

Feature selection is an important topic in ML because good predictor requires many features that are able to predict unique aspects of the outcome variable. 

## Research question and data

Our research question is:  

**What factors are associated with BMI?**

We have already worked on the imputed data from the missing data class so we are going to use that imputed data here as a starting point for the work. This is partly to save time for the teaching part of the class. Remember from the missing data class that ideally we would

1. Conduct sensitivity analyses with a number of different imputation methods. 
2. Account for uncertainty using pooled models. 

We are going to skip that for now but I will show an example of how `tidymodels` imputes data later in this doc. 

#### Outcome variable

Let's look at the outcome variable, recode, and drop observations that are not relevant. We need to do a histogram and check the distribution. Then we might deal with outliers.  

```{r}
summary(data$PM_BMI_SR)

bmi_histogram <- ggplot(data = data, aes(PM_BMI_SR)) +
                  geom_histogram(binwidth = 2)
plot(bmi_histogram)
```

Nice normal(ish) distribution here. We probably have some outliers on the low and high end with values of 8.86 and 69.40 

We can recode people who are less than 10 and greater than 60 to values of 10 and 60 respectively. 

```{r}
data <- data %>%
          mutate(bmi_recode = case_when(
            PM_BMI_SR < 10 ~ 10, 
            PM_BMI_SR > 60 ~ 60,
            TRUE ~ PM_BMI_SR
          ))
summary(data$bmi_recode)

bmi_recode_histogram <- ggplot(data = data, aes(bmi_recode)) +
                  geom_histogram(binwidth = 2)
plot(bmi_recode_histogram)
```

### Preparing predictor variables

All of the predictors are coded as 0,1,2 are read in as numeric by R so we need to fix that. We could manually fix each variable but we are going to do something a bit different. All of the `_EVER` variables are coded as

    * 0 Never had disease
    * 1 Ever had disease
    * 2 Presumed - Never had disease
    * -7 Not Applicable

We can batch recode all of these variables and make sure that they are factor and not numeric.

```{r}
data <- data %>% mutate_at(3, factor)
data <- data %>% mutate_at(5:6, factor)
data <- data %>% mutate_at(8:12, factor)
data <- data %>% mutate_at(15:81, factor)
data <- data %>% mutate_at(83:93, factor)
```

## Recoding

**Age**

```{r}
summary(data$SDC_AGE_CALC) 

### Checking NA

data %>% summarise(
                  n = n_distinct(SDC_AGE_CALC),
                  na = sum(is.na(SDC_AGE_CALC)
                           ))

```

**Income**

```{r}
glimpse(data$SDC_INCOME)

table(data$SDC_INCOME)

data <- data %>%
	mutate(income_recode = case_when(
		SDC_INCOME == 1 ~ "Less than 25 000 $",
		SDC_INCOME == 2 ~ "Less than 25 000 $",
		SDC_INCOME == 3 ~ "25 000 $ - 49 999 $",
		SDC_INCOME == 4 ~ "50 000 $ - 74 999 $",
		SDC_INCOME == 5 ~ "75 000 $ - 99 999 $",
		SDC_INCOME == 6 ~ "100 000 $ - 149 999 $",		
		SDC_INCOME == 7 ~ "150 000 $ - 199 999 $",
		SDC_INCOME == 8 ~ "200 000 $ or more"
	))

glimpse(data$income_recode)

data$income_recode <- as_factor(data$income_recode)

data$income_recode <- fct_relevel(data$income_recode, "Less than 25 000 $", 
                                                          "25 000 $ - 49 999 $",
                                                          "50 000 $ - 74 999 $",
                                                          "75 000 $ - 99 999 $",
                                                          "100 000 $ - 149 999 $",
                                                          "150 000 $ - 199 999 $",
                                                          "200 000 $ or more"
                                          )
table(data$income_recode)

table(data$income_recode, data$SDC_INCOME)
```

### Some final data organizing

```{r}
data$PM_BMI_SR <- NULL
```

We have run basic model with all variables but maybe that's too slow and maybe it's overfitting the data and we want to use a different approach for model selection. We can use ridge and lasso regression (as discussed in the slides). We already have our workflow and recipe from tidymodels so all we need to change is your regression part of the tidymodels section and we run a ridge regression. We can't use `glm` for the this as it does not support these methods. We need to change the engine to `glmnet`. 

### BMI recipe

```{r}
# Fix the random numbers by setting the seed 
# This enables the analysis to be reproducible when random numbers are used 
set.seed(10)

data_split <- initial_split(data, prop = 0.70)

# Create data frames for the two sets:
train_data <- training(data_split)
summary(train_data$bmi_recode)

test_data  <- testing(data_split)
summary(test_data$bmi_recode)

bmi_recipe <- recipe(bmi_recode ~ ., data = train_data) %>% 
  update_role(ADM_STUDY_ID, new_role = "ID") %>% 
  update_role(ID, new_role = "ID") %>% 
  step_normalize(all_numeric_predictors()) %>% 
  step_dummy(all_nominal_predictors()) %>% 
  step_zv(all_predictors())

summary(bmi_recipe)
```

The tuning parameter for the ridge regression is known has a __hyperparameter__. Different types of ML models will have different hyperparameters that you need to tune to make sure that they are optimized for your model. 

Here sort of randomly set the penalty for the ridge model but that's just to give us an example. We will tune the penalty a bit later. The code for this is mostly based from [https://juliasilge.com/blog/lasso-the-office/](https://juliasilge.com/blog/lasso-the-office/) this tutorial, with some modifications for explanability and because some functions are depricated. 


```{r}
ridge_model <- linear_reg(penalty = 0.1, mixture = 0) %>%  ### mixture = 0 is ridge, mixture = 1 is lasso. 
  set_engine("glmnet")

ridge_workflow <- workflow() %>%
  add_recipe(bmi_recipe) 

ridge_fit <- ridge_workflow %>%
  add_model(ridge_model) %>%
  fit(data = train_data)

ridge_fit %>%
        extract_fit_parsnip() %>%
        tidy()
```

### Tunning the penalty. 

So we fit one lasso model, but how do we know the right regularization parameter penalty? We can figure that out using resampling and tuning the model. Let’s build a set of bootstrap resamples, and set penalty = tune() instead of to a single value. We can use a function penalty() to set up an appropriate grid for this kind of regularization model.

First, we need to setup the bootstrap for the training data to be able to easily run multiple iterations of the model. Here instead of giving a fixed penalty value `penalty = 0.1` like we did above. We are going to get that value to `tune()` so we can run a number of different iterations of the penatly value and see which one is best. We are going to use a regular grid for this with 4 different iterations of the tuning. Normally, you would run more than 4 iterations but just showing as an example that runs relatively quickly. 

```{r}
set.seed(123)
data_boot <- bootstraps(train_data, times = 4)

lm_mixture <- linear_reg(
                mode = "regression",
                engine = "glmnet",
                penalty = tune(),
                mixture = tune()
              )
```

Here we are going to run the model 4 times with different penalty values. That's computationally expensive so let's do this in parallel. 

```{r}
### This takes around 45minutes to run on my 8core computer

detectCores()
doParallel::registerDoParallel()

set.seed(123)

bmi_wflow_boots <- workflow() %>% 
          add_model(lm_mixture) %>% 
          add_recipe(bmi_recipe) %>% 
          tune_grid(resamples = data_boot,
                    control = control_grid(save_pred = TRUE, 
                                            verbose = FALSE)) ## Edit for running live

collect_metrics(bmi_wflow_boots) 

show_best(bmi_wflow_boots, metric='rmse', n=5)  # only show the results for the best 5 models
```

### Plotting

```{r}
autoplot(bmi_wflow_boots, metric = 'rmse') 

bmi_wflow_boots %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  mutate(penalty = factor(penalty)) %>%
  ggplot(aes(mixture, mean, color = mixture)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point() +
  labs(y = "Root Mean Square Error")
```

```{r}
ridge_lasso_best <- 
  bmi_wflow_boots %>% 
  select_best(metric = "rmse")

ridge_lasso_best
```

Let's pull the results and see what we get

```{r}
boot_tuned_results <- 
  bmi_wflow_boots %>% 
  collect_predictions(parameters = ridge_lasso_best) 
```

### Final Model

Final model

Above we are looking at our trained model over the cross-validation sets. We have not actually tested our model on our test data. To run the last model we need to back to our workflow and use the last_fit function. Note that we use the cv_split object rather than the train or test data objects. This will will fit the model to the entire training set and evaluate it with the testing set. We need to back to our workflow object (somewhat counter intuitive).

```{r}
lm_mixture_best <- linear_reg(
                mode = "regression",
                engine = "glmnet",
                penalty = 0.0774,
                mixture = 0.261
              )

bmi_wflow_best <- workflow() %>% 
          add_model(lm_mixture_best) %>% 
          add_recipe(bmi_recipe) 


final_ridge_model_workflow <- finalize_workflow(
                    bmi_wflow_best,
                    ridge_lasso_best
                  )

final_ridge_model_workflow
```

### Variable importance 

```{r}
final_ridge_model <- final_ridge_model_workflow %>%
                        fit(data = train_data)

final_ridge_model %>%
        extract_fit_parsnip() %>%
        tidy()

vip(final_ridge_model, num_features = 20)
```

### Final model on testing data 

let’s return to our test data. The tune package has a function last_fit() which is nice for situations when you have tuned and finalized a model or workflow and want to fit it one last time on your training data and evaluate it on your testing data. You only have to pass this function your finalized model/workflow and your split. As we pass it the split object from the start of our code we get two processes for the price of one. We train/fit our selected model on 100% of Train and then it automatically scores up the Test set with the newly created, final model. 

```{r}
last_fit_ridge <- last_fit(final_ridge_model_workflow, data_split)  %>%
                    collect_metrics()
last_fit_ridge
```

## Session Info

```{r}
sessionInfo()
```

