---
title: "Validation Set"
author: "Daniel Fuller"
output:
      html_document:
        keep_md: true
---

```{r setup, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidymodels)
library(sjPlot)
library(finalfit)
library(knitr)
library(gtsummary)
library(mlbench)
library(kernlab)
library(vip)
library(rsample)
library(tune)
library(recipes)
library(yardstick)
library(parsnip)
library(glmnet)
library(themis)
library(microbenchmark)
```

# Validation Set

## Research question and data

We are using an imputed (ie. no missing data) version of the CanPath student dataset [https://canpath.ca/student-dataset/](https://canpath.ca/student-dataset/). The nice thing about this dataset is that it's pretty big in terms of sample size, has lots of variables, and we can use it for free. 

Our research question is:  

- **Can we develop a model that will predict type 2 diabetes**

## Validation Set

Here we rather than using multiple iterations of resampling, let’s create a single resample called a validation set. In tidymodels, a validation set is treated as a single iteration of resampling. We’ll use the validation_split() function to allocate 20% of the data to the validation set and 10% to the training set. This means that our model performance metrics will be computed on the single validation set. This is fairly large, so the amount of data should provide enough precision to be a reliable indicator for how well each model predicts the outcome with a single iteration of resampling.

![](https://www.tidymodels.org/start/case-study/img/validation-split.svg)

### Reading in data

Here are reading in data and getting organized to run our models. 

```{r}
data <- read_csv("canpath_imputed.csv")

data <- data %>% mutate_at(3, factor)
data <- data %>% mutate_at(5:6, factor)
data <- data %>% mutate_at(8:9, factor)
data <- data %>% mutate_at(12:12, factor)
data <- data %>% mutate_at(15:81, factor)
data <- data %>% mutate_at(83:93, factor)

table(data$DIS_DIAB_EVER)

data <- data %>%
	mutate(diabetes = case_when(
		DIS_DIAB_EVER == 0 ~ 0,
		DIS_DIAB_EVER == 1 ~ 1,
		DIS_DIAB_EVER == 2 ~ 0)) %>%
		mutate(diabetes = as.factor(diabetes))

table(data$DIS_DIAB_EVER, data$diabetes)

data$DIS_DIAB_EVER <- NULL
```

```{r}
data <- select(data, diabetes, 
                            PSE_ADULT_WRK_DURATION, 
                            PM_BMI_SR, 
                            PA_TOTAL_SHORT, 
                            SDC_HOUSEHOLD_CHILDREN_NB, 
                            SDC_HOUSEHOLD_ADULTS_NB, 
                            SDC_EDU_LEVEL_AGE, 
                            SDC_AGE_CALC, 
                            SDC_GENDER, 
                            SDC_INCOME)
```

## Prepare the data split and cross validation folds

This works across all models so we only need to run this once. 

```{r}
#### Cross Validation Split
cv_split <- initial_validation_split(data, 
                            strata = diabetes, 
                            prop = c(0.70, 0.20))

# Create data frames for the two sets:
train_data <- training(cv_split)
table(train_data$diabetes)

validation_data  <- validation(cv_split)
table(validation_data$diabetes)

test_data  <- testing(cv_split)
table(test_data$diabetes)
```

### V folds

```{r}
folds <- vfold_cv(training(cv_split), v = 5, strata = diabetes)
```

## Prepare the recipe for the data 

This works across all models so we only need to run this once. 

```{r}
diabetes_recipe <- 
  recipe(diabetes ~ ., data = data) %>%
  step_smotenc(diabetes, over_ratio = 0.9) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors(), -all_outcomes()) %>%
  step_normalize(all_numeric_predictors())

diabetes_recipe_validation <- 
  recipe(diabetes ~ ., data = validation_data) %>%
  step_smotenc(diabetes, over_ratio = 0.9) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors(), -all_outcomes()) %>%
  step_normalize(all_numeric_predictors())
```

## Prepare the models 

Here we need to understand the tuning parameters, which are different for each type of model and specify which parameters we will run and how we will run the grid search for tuning. I'm going to do this in one chunk for each just to make it easier to follow. 

### Logistic regression

Tuning parameters

* __mixture__: A number between zero and one (inclusive) giving the proportion of L1 regularization (i.e. lasso) in the model.
    * mixture = 1 specifies a pure lasso model
    * mixture = 0 specifies a ridge regression model
    * ⁠0 < mixture < 1⁠ specifies an elastic net model, interpolating lasso and ridge
* __penalty__: A non-negative number representing the total amount of regularization (specific engines only). For keras models, this corresponds to purely L2 regularization (aka weight decay) while the other models can be either or a combination of L1 and L2 (depending on the value of mixture).


```{r Logistic Training}
logistic_model <- logistic_reg(penalty = tune(), mixture = tune(),
                                mode = "classification",
                                engine = "glmnet"
                               )

logistic_workflow_train <- workflow() %>% 
          add_model(logistic_model) %>% 
          add_recipe(diabetes_recipe) %>% 
          tune_grid(train_data,
                    resamples = folds,
                    control = control_grid(save_pred = TRUE, 
                                            verbose = FALSE)) ## Edit for running live

collect_metrics(logistic_workflow_train) 

show_best(logistic_workflow_train, metric='brier_class', n=5)  # only show the results for the best 5 models

autoplot(logistic_workflow_train, metric = 'brier_class') 
```

## Validation Set

```{r Logistic Validation}
logistic_workflow_validation <- workflow() %>% 
          add_model(logistic_model) %>% 
          add_recipe(diabetes_recipe) %>% 
          tune_grid(validation_data,
                    resamples = folds,
                    control = control_grid(save_pred = TRUE, 
                                            verbose = FALSE)) ## Edit for running live

collect_metrics(logistic_workflow_validation) 

show_best(logistic_workflow_validation, metric='brier_class', n=5)  # only show the results for the best 5 models

autoplot(logistic_workflow_validation, metric = 'brier_class') 
```

### Compare testing and validation

```{r}
logistic_best_train <- 
  logistic_workflow_train %>% 
  select_best(metric = "brier_class")

logistic_best_validation <- 
  logistic_workflow_validation %>% 
  select_best(metric = "brier_class")

logistic_best_train
logistic_best_validation
```

```{r Logistic Final Model}
logistic_best <- 
  logistic_workflow_validation %>% 
  select_best(metric = "brier_class")

logistic_final_model <- finalize_model(
                          logistic_model,
                          logistic_best
                          )
logistic_final_model

final_logistic_workflow <- workflow() %>%
                      add_recipe(diabetes_recipe) %>%
                      add_model(logistic_final_model)

final_logistic_results <- final_logistic_workflow %>%
                    last_fit(cv_split)

lr_results <- final_logistic_results %>% collect_metrics()
```


## Final Results

```{r}
### Logistic Regression
kable(lr_results)
```

## Variable Importance

### Logistic Regression

```{r}
prep <- prep(diabetes_recipe)

logistic_final_model %>%
  set_engine("glmnet", importance = "permutation") %>%
  fit(diabetes ~ .,
    data = juice(prep)) %>%
  vip(geom = "point")
```

## Session Info

```{r}
sessionInfo()
```

